\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
\usepackage{authblk}
%\geometry{landscape}                		% Activate for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex
								
%\usepackage{{Gratzer-Color-Scheme}	% Active to color theorems red and lemmas blue
	
\usepackage{amssymb}
\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers

\usepackage{natbib}
\setcitestyle{aysep={}} 
\usepackage{hyperref}

%SetFonts

%SetFonts

\date{}							% Activate to display a given date or no date

\begin{document}
\title{The Future of MODFLOW -- post 2025}

\author[1]{Michael N. Fienen}
\affil[1]{\small U.S. Geological Survey, Upper Midwest Water Science Center, Madison, WI, USA}
\author[2]{Joseph D. Hughes}
\affil[2]{\small INTERA Incorporated, 927 W Belle Plaine Ave, Chicago, IL, USA}
\author[3]{Christian D. Langevin}
\affil[3]{\small S.S. Papadopulos \& Associates, Inc., 12 Dove Lane, Saint Paul, MN, USA}
\author[4, *]{Eric D. Morway}
\affil[4]{\small U.S. Geological Survey, Nevada Water Science Center, Carson City, NV, USA}
\author[5]{Alden M. Provost}
\affil[5]{\small U.S. Geological Survey, Integrated Modeling and Prediction Division, Reston, VA, USA}
\author[6]{Martijn J. Russcher}
\affil[5]{\small Deltares, Delft, NL}
\affil[*]{\small Corresponding author \href{mainto:emorway@usgs.gov}{emorway@usgs.gov}}

\maketitle

The MODFLOW groundwater modeling software code was originally developed by the United States Geological Survey (USGS) in the early 1980s to create a single program that combined the best capabilities of all the variations of groundwater model codes used in USGS studies \citep{mcdonald2003}. Major releases of early versions of MODFLOW were developed internally by the USGS and released approximately every 4 to 8 years \citep{modflow84, modflow88, harbaugh1996user, hill2000modflow, modflow2005}. Eventually, capabilities developed by non-USGS scientists, in collaboration with the USGS, began to be added to MODFLOW \citep[for example,][]{hoffmann2003modflow, wilson2004modflow, bakker2013documentation} or MODFLOW variants \citep[for example,][]{modflowvsf, modflownwt, modflowusg}. %SHOULD NAME CHECK OWHM and/or SURFACT HERE?

MODFLOW 6, the latest version of MODFLOW, was released in 2017 and consolidated popular capabilities available in MODFLOW variants into a single code \citep{modflow6gwf}. MODFLOW capabilities were quickly extended with the addition of groundwater transport \citep{modflow6gwt}, groundwater energy \citep{morway2025new}, and particle tracking models. Unlike previous versions of MODFLOW, modern open-source software development approaches and direct collaboration with external partners (for example, Deltares, University Corporation for Atmospheric Research, Delft University of Technology, etc.\@) has accelerated feature development. These approaches also enable a common core to advance collaboratively with the ability to develop separate capabilities in branches that inherit the core. This differs from previous codes like MODFLOW-OWHM and MODFLOW-SURFACT (REFS) that were added staticly to previous MODFLOW versions. Once that hard fork was made, the connection was broken. Maintaining this common core, collaboratively, is a key modern software design precept and the highest priority to ensure longevity and innovation of MODFLOW going forward.

As of 2026, the MODFLOW core now supports incorporation of additional hydrologic processes, allows interaction between processes, multiple model instances, multiple types of models within a single simulation, parallel simulation capabilities, and interoperability that allows MODFLOW to be accessed and controlled by external programs \citep{langevin2024modflow}. 

In 2025, there was approximately a 20\% reduction in USGS staff due to the new administration's initiatives and the Deferred Resignation Program. Included in this reduction were several key MODFLOW developers who left the USGS for private industry. Naturally, this led to many MODFLOW users and skeptics wondering if MODFLOW development and support would continue in the future. We regrouped To address this, the USGS and core MODFLOW contributors developed a \href{https://github.com/MODFLOW-ORG/modflow6/wiki/MODFLOW-Governance-Charter}{MODFLOW Governance Charter} to further broaden the development and decision-making base to formally include a broader cross-section of partners from the private and academic sectors of the MODFLOW community, particularly at the technical level. The Governance enables unfettered innovation in a way that also keeps the MODFLOW core functional for all needs, mitigating technical conflicts as they arise.

This charter defines MODFLOW as an open-source core framework for government, private, and academic use. While the USGS retains final authority over formal releases and high-level strategy, the project is technically guided by a Technical Steering Committee (TSC).
The TSC oversees software improvements, new features, and documentation. It consists of at least two USGS developers and two community members who serve renewable three-year terms. Membership and document changes are managed through a majority vote, ensuring a collaborative approach to the software's long-term maintenance. TSC members are required to have 1) a track record of contributing to MODFLOW development, 2) advanced technical skills directly relevant to the core software's development, such as proficiency in the underlying model framework or related technical functions, and 3) must show an ongoing ability and willingness to participate in both day-to-day development and high-level decision-making. The current TSC membership, also the authorship of this Editorial, is listed at \href{https://github.com/MODFLOW-ORG/modflow6/wiki/Technical-Steering-Committee-Meetings}{MODFLOW TSC Members}.

With USGS management approval of and support for the charter, we are confident that MODFLOW development and maintenance will continue. Furthermore, the charter provides a defined mechanism for continued development and improvement in addition to succession of new TSC members when current members resign, are voted out, or retire. 

2025 was a tumultuous year for MODFLOW, but the path forward is strong. The TSC provides continuity, with ongoing development from the core team while providing new opportunities for community involvement and direction. The result will be a core MODFLOW code that serves many different needs and is improved intentionally and robustly into the future.

%TC:ignore
\noindent\textbf{554/675 words}
%TC:endignore 

\bibliographystyle{groundwater}
\bibliography{modflow-editorial}


\end{document}  